## STAT 545A Final Project: Explore the temporal and spatial distribution of Canadian Large Fire
### Yumian Hu
__2013-10-21__

###  <dim id="1a">[Introduction](#1a)
In this project, I will use the **Canadian Large Fire Database (LFDB)** to explore the temporal and spatial distribution and association of landscape-scale fire in Canada. LFDB is a compilation of forest fire data from all Canadian agencies, including provinces, territories, and Parks Canada. The data set includes only fires greater than 200 hectares in final size; these represent only a few percent of all fires but account for most of the area burned (usually more than 97%). For more information, please visit website of [Natural Resources Canada](http://cwfis.cfs.nrcan.gc.ca/en_CA/lfdb), LFDB can also be downloaded ditrectly from this webpage.

### <dim id="2a">[Big Picture](#2a)
In thie project, I will explore from the following aspects.
* <dim id="3a">[Data Description and Import](#3b)
* <dim id="4a">[Temporal Distribution of Large Fire](#4b)
  * <dim id="4.1a">[How is Fire Frequency Change over Year](#4.1b)
  * <dim id="4.2a">[How is Fire Size Change over Month](#4.2b)
* <dim id="5a">[Spatial Distribution of Large Fire](#5b)
  * <dim id="5.1a">[Fire Frequency seen from Google Map](#5.1b)
  * <dim id="5.2a">[Emulate Real Map with `ggplot2`](#5.2b) 
* <dim id="6a">[Seen from both Temporal and Spatial Aspects --- Bubble Chart](#6b)
* <dim id="7a">[Heatmap: Is the monthly distribution of Fire Frequency same Every Year? ](#7b)
* <dim id="8a">[Fluctuation of Fire Frequency over Year in Different Provinces](#8b)


### <dim id="3b">[Data Description and Import](#3a)
The dataset is comma-delimited ASCII, named "LFD_A02_5999_e.txt" originally, with each row representing a record for a given fire. 
The raw data incorporates the following 16 variables: 
* `Year`, `Month`, `Day`, `Province`, `Fire_ID`, `Latitude`, `Longitude`, `Start_Date`, `Detect_Date`, `Cause`, `Size`, `Fire_Region`, `Fire_Zone`, `EcoZone`, `EcoRegion`, `EcoDistrict`. 

I include 8 most meaningful variables in my data analysis, and summarize thier information and description in the table below.

```{r echo = FALSE, results = 'asis'}
library(xtable)
DataDes <- read.table("result/DataFormat.tsv", header = TRUE)
print(xtable(DataDes), type = "html", include.rownames = TRUE)
```
> There is no record of January in the dataset, quite strange.
 
* One needs to sepecify the arguments `quot="\""` and `comment.char=""` when importing the raw data, since there are symbols of `'` and `#` in the dataset. 
* I assume most of the missing data is missing completely at random, which can be directly excluded from the dataset. However, I keep "UNK" Fire Causes (entry of missing value for `Cause`) in my clean dataset since its percentage is relatively large and it seems to have some missing pattern (all the fire causes in BC in 1999 were unkown!). 
* I have already excluded two levels of Province, WBNP (Wood Buffalo National Park) and NP (National Park) from the raw data, which I think hard to make comparison with "real"" provinces.
* Please visit [Wikipedia](http://en.wikipedia.org/wiki/Canadian_postal_abbreviations_for_provinces_and_territories) for more information about Canadian provinces and their abbreviations. Since the Large Fire Database is not updated, it incorporate some old versions of abbreviations, I update them in the clean dataset with "PQ", "NWT", "NF", "YK" changed to "QC", "NT", "NL" and "YT".
* There are 11231 rows and 16 column in the raw dataset. After some data cleaning and maniputaion, there are 10874 rows of 8 columns left in the clean dataset.

Now let's import the clean dataset and load necessary packages.
```{r include = FALSE}
opts_chunk$set(tidy = F, warning=F, comment=NA, prompt=T)
```

```{r}
jDat <- read.table("data/CleanData.tsv", header=T)
```

```{r}
library(plyr)          # for data aggregation
library(RColorBrewer)  # for color setting
library(ggplot2)       # for advanced figure plotting 
library(gplots)        # for heatmap
library(RgoogleMaps)   # for showing map of Canada

```

Perform a superficial check that data import went OK.
```{r}
str(jDat)
head(jDat)
```

Reorder the levels of Province: From West to East and from South to North and sort the data by Year and Month
```{r}
ProLevels <- c("BC", "AB", "SK", "MB", "ON", "QC", "NB", "NS", "NL", "YT", "NT")
jDat <- arrange(transform(jDat, Province = factor(Province, levels = ProLevels)), Province)
jDat <- jDat[with(jDat, order(Year, Month, Day, Province, Size)), ]
str(jDat)
```

### <dim id="4b">[Temporal Distribution of Large Fire](#4a)
As a basic check, let's first look at how Fire Frequency and Fire Size change with time.

#### <dim id="4.1b">[How is Fire Frequency Change over Year](#4.1a)
```{r echo = FALSE, fig.show='hold', out.width='50%'}
gDat <- ddply(jDat, ~ Year, function(x) c(TotalFire = length(x$Year)))
ggplot(gDat, aes(x = Year, y = TotalFire)) + geom_point() + geom_line() + 
  theme_bw() + ggtitle("Total Number of Large Fire over Year")
gDat <- ddply(jDat, ~ Year + Cause, function(x) c(TotalFire = length(x$Year)))
ggplot(gDat, aes(x = Year, y = TotalFire, color = Cause)) + 
  geom_point() + geom_line() + theme_bw() +
  ggtitle("Total Number of Large Fire over Year by Different Causes")
```

I plot the total number of Large Fire against Year, and grouped by Causes in the right figure. There is no obvious increasing or decreasing trend for fire frequency. It's rather fluctuated with its peak at the year of 1979. Most of the fire, of course, was caused by lightning but still rather amount was caused by human, especially from 1959 to 1969.

#### <dim id="4.2b">[How is Fire Size Change over Month](#4.2a)
```{r, echo = FALSE, fig.width=10, fig.height=9}
jColors <- brewer.pal(n = length(unique(jDat$Province)), name = "Set3")
names(jColors) <- unique(jDat$Province)

jYear <- seq(from = 1959, to = 1999, by = 10)

ggplot(subset(jDat, Year %in% jYear), 
       aes(x = factor(Month), y = Size, color = Province)) + 
  scale_y_log10() + geom_point(alpha = I(0.2)) + geom_jitter() + 
  facet_wrap(~ Province, nrow = 4) + theme_bw() + 
  scale_color_manual(values = jColors) + 
  xlab("Month") + ylab("Fire Size") +
  ggtitle("Fire Size over Month in Different Province")
```
To make the above figure, I choose records from year 1959, 1969, 1979, 1989 and 1999, plot the log transformed fire size against its happening month and wrap by province. Most of the provinces show the similar volcano shape, centered from May to August, which implies that Summer and Autumn is the frequent seasons for large fire. 

### <dim id="5b">[Spatial Distribution of Large Fire](#5a)
Let's move to the spatial part, look at the geographically distribution of Large Fire location every 5 years from 1949 to 1999.

#### <dim id="5.1b">[Fire Frequency seen from Google Map](#5.1a)
Since the dataset incorporate Latitude and Longitude of each Fire, I can download the GoogleMap of Canada and plot the fire location in this map. You need to install `RgoogleMaps` package first. 
```{r}
center = c(mean(jDat$Latitude), mean(jDat$Longitude))
zoom <- min(MaxZoom(range(jDat$Latitude), range(jDat$Longitude)))
GeoMap <- GetMap(center = center, zoom = zoom, destfile = "figure/capre.png")
jYear <- seq(from = 1959, to = 1999, by = 5)
gDat <- subset(jDat, Year %in% jYear)
PlotOnStaticMap(GeoMap, lat = gDat$Latitude, lon = gDat$Longitude, 
                destfile ="figure/capre.png", cex = 0.5, pch = 20, col = "red")
```

#### <dim id="5.2b">[Emulate Real Map with `ggplot2`](#5.2a) 
I would like to emulate it with `ggplot2` and distinguish Month by different colors. 
```{r, echo = FALSE, fig.width=9, fig.height=9}
gColors <- brewer.pal(n = length(unique(jDat$Month)), name = "Set3")
names(gColors) <- sort(unique(jDat$Month))
ggplot(gDat, aes(x = Longitude, y = Latitude, color = factor(Month))) + 
  geom_point(alpha = I(0.8)) + theme_bw() + 
  scale_color_manual(name = "Month", values = gColors) +
  ggtitle("Geographical Distribution of Large Fire")

```
Again it shows the seasonal trendence with the most Fire happening from May to August and in the middle of Canada (SK, MB, ON). 

### <dim id="6b">[Seen from both Temporal and Spatial Aspects --- Bubble Chart](#6a)
In this part, I will replicate the Bubble Chart from Gapminder Project, which displays both the Temporal and Spatial trends of Large Fire. Based on the above figure, I make the area of solid circle proportional to its fire size at that location (not the exact fire size if it exceeds its provincial territory).

```{r, echo = FALSE, fig.width=9, fig.height=9}
jDarkGray <- 'grey20'
ggplot(subset(jDat, Year %in% jYear), 
       aes(x = Longitude, y = Latitude, fill = factor(Month))) + 
  geom_point(aes(size = sqrt(Size/pi)), pch = 21, color = jDarkGray) +
  scale_size_continuous(range=c(1,40)) + theme_bw() +  
  facet_wrap(~ Year) + 
  scale_fill_manual(values = gColors) +
  guides(size = FALSE) +
  ggtitle("Geographical Distribution of Large Fire by Month with Area Proportional to Fire Size")
```
* Some super large fire happened in July, so there are not only frequent fire but also large fire happening in summer.
* There was a huge fire happened in the area around YT and NT in 1979. But there is no fixed pattern of geographical distribution for large fire. 

### <dim id="7b">[Heatmap: Is the monthly distribution of Fire Frequency same Every Year? ](#7a)
To better show the temporal distribution of fire, I use heatmap to demonstrate the number of fire happening at every month and year. The darker the blue is, the more fire it presents. 
```{r}
gDat <- daply(jDat, ~ Year + Month, function(x) length(x$Year))
## replace NA with 0 in the gDat
gDat[is.na(gDat)] <- 0
gDat <- as.data.frame(gDat)
colnames(gDat) <- paste0("Month ",colnames(gDat))
rownames(gDat) <- paste0("Year ", rownames(gDat))
gDat <- as.matrix(t(gDat))
gBuPuFun <- colorRampPalette(brewer.pal(n = 9, "BuPu"))
heatmap.2(gDat, col = gBuPuFun, trace = "none", main = "Heatmap of the Number of Fire")
```

To explore whether the monthly distribution of fire frequency similar in  every year, I first calculate the correlation between every two years and then show the result with heatmap.
```{r}
gDatCor <- cor(gDat)
heatmap.2(gDatCor, col = gBuPuFun, trace = "none", 
          main = "Heatmap of Correlation among Years")
```

It shows that the color of grid between year 1960 and 1995 is the lightest, which indicates the lowest correlation. We can also demonstrate it from the correlation matrix.

```{r}
min(gDatCor) 
minrow <- which.min(gDatCor) %/% nrow(gDatCor) + 1
mincol <- which.min(gDatCor) %% nrow(gDatCor)
gDatCor[minrow, mincol] == min(gDatCor)
(year1 <- as.numeric(substr(rownames(gDatCor)[minrow], 6, 9)))
(year2 <- as.numeric(substr(colnames(gDatCor)[mincol], 6, 9)))
```

Let's plot the monthly trend of fire frequency for each year and highlight the year 1960 and 1995.
```{r, echo = FALSE}
gDat <- ddply(jDat, ~ Year + Month, function(x) c(TotalFire = length(x$Year)))
## The tow years are plotted with red and blue, others are grey
PickYr <- range(minrow, mincol)
gColors <- rep(c("grey", "deeppink", "grey", "red", "grey"), 
               c(PickYr[1]-1, 1, PickYr[2]-PickYr[1]-1, 1, nrow(gDatCor)-PickYr[2]))
names(gColors) <- sort(unique(gDat$Year))

ggplot(gDat, aes(x = Month, y = TotalFire, color = factor(Year))) + 
  geom_point() + geom_line() + theme_bw() + 
  scale_color_manual(values = gColors) +
  ggtitle(paste0("Total Number of Fire over Month in year ", year2, ", ", year1)) +
  guides(color = FALSE) 
```

### <dim id="8b">[Fluctuation of Fire Frequency over Year in Different Provinces](#8a)
The following figure displays the fire frequency trend over year in each province. Most of them are quite fluctuated which makes it meaningless to fit a linear regression line. However, we can still compare the fluctuation of fire frequency among different provinde. 

```{r, fig.width=9, fig.height=9}
gDat <- ddply(jDat, ~ Year + Province + Cause, function(x) c(TotalFire = length(x$Year)))
gColors <- brewer.pal(n = nlevels(gDat$Cause), name = "Dark2")
names(gColors) <- sort(levels(gDat$Cause))

ggplot(gDat, aes(x = Year, y = TotalFire, color = Cause)) + 
  geom_point() + geom_line() + theme_bw() + 
  facet_wrap(~ Province, nrow=4) + 
  scale_color_manual(values = gColors) +
  ggtitle("Total Number of Large Fire over Year in Different Provinces")
```

I first fit a cubic spline (A cubic spline is a spline constructed of piecewise third-order polynomials which pass through a set of control points) for each type of fire cause per province. Then I use the average curvature (curvature/length of year) as a representative of fluctuation of trend; the larger the average curvature is, the more fluctuation the curve has. 
> I drop the province of "NB" and "NS" since they don't have enough number of points (at least four) to fit the cubic spline.

```{r echo = FALSE, results = 'asis'}
## exlude  province of "NB", "NS" and cause of "UNK"
source("code/SplineCurv.R")
hDat <- droplevels(subset(gDat, Cause != "UNK" & ! Province %in% c("NB", "NS")))
hDat <- daply(hDat, ~ Province + Cause, function(x) 
  {my.spline(x$Year, x$TotalFire)/length(x$Year)})
hDat <- as.data.frame(hDat, include.rownames = T)
hDat <- hDat[with(hDat, order(-LTG)), ]
print(xtable(hDat), type = "html", include.rownames = TRUE)
```
Therefore, the fire frequency trend caused by lightning fluctuated the most in NT and the ond caused by human fluctuated the most in MB.

